\chapter{Background}\label{chap:background}

\section{The Nanopass Compiler}

The compiler's architecture is based on Jeremy Siek's textbook \textit{Essentials of Compilation}~\citep{siek2023essentials}. The main idea is that instead of performing one huge transformation, the compilation process is separated into a sequence of passes, each with its own target representation and task.

The first step is parsing the input program into an AST. Several third-party tools and libraries exist for parsing, and building one is not the scope of this thesis; this compiler relies on Python's own \texttt{ast} module~\citep{pythonastdocs}. Once a Python AST is available, each pass has a designated and well-specified task, leading to a clean architecture in which each module has its own job~\citep{martin2008cleancode}.

Among these passes are, for example, flattening out functions, bringing a nested expression into a list of atomic expressions (monadic normal form), selecting instructions, and allocating addresses and space on the stack or heap.

While the textbook and Professor Thimann's lecture cover many concepts of a compiler, support for floating-point values and their operations was not present in the textbook or lecture version of the compiler. During this thesis, I explored how the existing architecture could be extended to support floating-point values, which design decisions had to be made, and what alternatives existed.

I ultimately arrived at an implementation of one such approach, which is discussed in the following chapters.

\section{RISC-V}

The target architecture of the compiler is the RISC-V instruction set architecture~\citep{waterman2016riscv}. RISC-V is an open instruction set architecture, originally developed at the University of California, Berkeley.

RISC-V's design philosophy favors a reduced number of supported instructions, making it a comparatively simple and abstract architecture to target from a compiler.

Unlike proprietary architectures, RISC-V is maintained as an open standard rather than by a single vendor, with multiple independent hardware and software implementations~\citep{waterman2016riscv}.


\section{IEEE 754 Floating Point Representation}

For encoding floating-point values, I decided to follow the industry standard, IEEE 754~\citep{ieee754-2019}. Under this standard, a double-precision float is represented as a 64-bit word consisting of 1 sign bit, an 11-bit exponent, and a 52-bit mantissa.

Double precision is natively supported by the RISC-V architecture and matches the precision Python uses internally to represent floats~\citep{pythonfloatdocs}. A problem emerged with the existing compiler design, in which the least significant bit (LSB) of every value and pointer was used for tagging - something that is not possible with this encoding, since it requires all 64 bits.

This was solved by boxing floats and encoding their type in the header of the box (see chapters 3 and 4).