\chapter{Implementation}\label{chap:implementation}
% The Engineering Gritty Details: How did you handle code generation for floating-point operations?
% Type Typing/Checking: How does your compiler catch type mismatches (e.g., trying to add an integer to a float if implicit casting isn't supported)?
% Runtime/Backend Integration: If there is a runtime system or specific assembly/IR emitted (like x86 SSE registers or LLVM float primitives), explain how you lowered the AST nodes to it.

The integration of floating-point operations requires a full-vertical extension across the compiler pipeline. This chapter details the type systems, lowering mechanisms, and code generation required to support 64-bit double-precision floats within the RV64GC architecture.

\section{Interpreter Extension and Phase-Driven Design}

Before implementing the concrete compiler backend, the language's dynamic semantics were established by extending the reference interpreter. Because the interpreter defines the ground-truth behavioral specification of the language, this execution layer must be finalized first to ensure semantic parity with the compiler's eventual target output.

To establish clear verification metrics, a differential test suite containing half a dozen Python programs was constructed. These test cases duplicated existing integer-based control flow, arithmetic expressions, and native built-in functions (such as \texttt{print()}), substituting integer literals with floating-point values. The immediate implementation goal was to allow the interpreter to execute this test suite natively without throwing exceptions, matching Python's runtime behavior exactly.

Extending the interpreter required systematic modifications across three core subsystems of the pipeline: the abstract syntax tree (AST) parser, the static type-checker, and the evaluation semantics engine.

\subsection{Syntactic Extensions (The Parser)}
Extending the frontend parser was highly deterministic because the host language's \texttt{ast} module natively recognizes floating-point tokens. The internal AST grammar definition was upgraded to allow literal expression nodes (\texttt{EConst}) to hold float primitives alongside integers. Additionally, the type-mapping routines were extended to recognize and propagate explicit variable type annotations, allowing patterns such as \texttt{a : float = 3.14} to pass successfully through the front-end token stream.

\subsection{Static Semantics and the 64-Bit Precision Decision}
The primary design challenge emerged during the static analysis phase within the type-checking subsystem. Because the compiler targets the RV64GC architecture—which natively supports both single-precision ('F') and double-precision ('D') floating-point extensions—implementing 32-bit floats offered tempting advantages in implementation expediency, specifically minimizing heap pressure and reducing the footprint of boxed values. However, forcing a 32-bit representation would introduce semantic drift from standard Python behavior, which treats all floating-point numbers as 64-bit doubles; common decimals like \texttt{0.1} or \texttt{3.14} would suffer silent precision loss. This directly conflicts with the design philosophy established in Chapter~\ref{chap:design_architecture}, which prioritises preserving Python's semantics over efficiency and implementation simplicity.

A hybrid architecture—selectively flagging exact base-2 powers (e.g., $0.5 = 2^{-1}$) as 32-bit floats and remaining decimals as 64-bit doubles—was also explored and rejected for the same reason: Python does not expose mixed-precision literals, and forcing such a distinction would create an un-Pythonic dialect while introducing significant complexity into the code generator.

To preserve strict semantic fidelity, I opted for a unified 64-bit double-precision model targeting the RISC-V 'D' extension, accepting the trade-off of higher memory overhead. Consequently, the type validator enforces a single constraint: verifying that literals do not exceed the 64-bit IEEE 754 infinity threshold.

\subsection{Type-System Generalization}
Beyond validating literals, the type-checker's expression evaluation routines needed a clean refactoring. Originally, the compiler assumed almost everything was an integer, evaluating binary operators and functions like \texttt{print()} by checking them directly against a hardcoded \texttt{TInt()} type.
To support floating-point numbers without copying and pasting the same validation logic everywhere, this rigid approach was replaced with a more flexible system. A new helper function, \texttt{check\_types\_supported}, was introduced to validate expressions against a list of acceptable types instead of just one. For this first iteration, operands must still match exactly—meaning mixed-type operations like \texttt{float + int} are blocked for now. However, this refactor ensures that expressions evaluate to and return their correct type primitive (like \texttt{TFloat()}) rather than defaulting to an integer, laying the clean groundwork for future cross-type arithmetic.

\subsection{Interpreter Semantics}
Extending the runtime behavior in the interpreter was the most straightforward step because the interpreter itself is written in Python, which already supports 64-bit floating-point math natively.

In \texttt{semantics.py}, the evaluation rules for binary expressions were simply updated to accept float values alongside integers. When the interpreter encounters an operation like addition, it handles the AST nodes by passing the raw values directly to Python's native operators. By leveraging the host language's execution environment this way, the interpreter was immediately able to run the floating-point test suite with perfect accuracy.

\section{Heap Allocation for Floats}

The decision to box every float carried two practical consequences for the allocation pass. From this pass onwards, a float is represented as a pointer to its box, so any operation that consumes a float must first unbox it to recover the underlying value, and any operation that produces a float must box the result again before it can be stored back on the heap. The difficulty is that the allocation pass must know, for each operand, whether it is a float; otherwise it would apply the arithmetic to the box pointer rather than to the value it points to. The type information needed to make this decision was computed earlier, by the type checker, so the task was to make that information available at this point in the pipeline and to consult it for each operand.

The first step was a function that determines whether a given expression yields a float. Whenever it does, the uniform boxing design dictates that the operand must be unboxed before the operation and, where the operation itself yields a float, boxed again afterwards.

A complication arose for values held inside tuples (and for closures, functions, and objects, which are all represented as tuples at this stage). The type checker does verify the type of every element of a tuple, but that per-element information was not preserved beyond type checking, so by the time the allocation pass ran there was no way to recover whether the element at a given position was a float. The expression's shape alone is sufficient for literals and for nested arithmetic, but not for accesses into a tuple or for plain variables, where the type is not visible in the expression itself. To resolve this, I returned to the type checker and preserved the type information it had already derived.

\subsection{A Dictionary to Preserve Type Information}

The result was a dictionary that, for each such structure, uses its identifier as the key and stores its fields or elements together with their corresponding types as the value. This dictionary is returned from the type checker to the main compiler driver, stored there, and passed on to every pass that requires it.

Up to this point in the pipeline, three passes consume the dictionary. The first is uniquify, which renames variables and must therefore update the dictionary's keys so that the recorded types remain reachable under the new names. The second is closure conversion, discussed below. The third is the heap allocator, for which the dictionary was originally introduced, where it is used to decide whether a given element is a float.

\subsection{Preserving Types Through Closure Conversion}

Closure conversion runs between uniquify and the heap allocator, and unlike uniquify it does not merely rename variables - it moves a lambda's free variables and parameters into an entirely new top-level function, generated on the fly under a fresh label. Any type information recorded under the enclosing function's identifier is therefore no longer reachable from within the new function's scope, precisely where the allocator would later need it to decide whether a captured variable or parameter is a float.

To keep this information reachable, closure conversion builds a small local type map for each lambda, restricted to its free variables and parameters, and registers it in the dictionary under the new function's own label, alongside the maps for the functions already present in the source program. This way, every function the allocator later encounters - whether written directly by the programmer or generated during closure conversion - has its own entry in the dictionary, keyed consistently by function identity.

\subsection{Handling Floats inside the Heap Allocator}

When one or more operands of an operation are floats, the allocator emits a block that first extracts each float value from its box, then performs the operation on the unboxed values, and finally, where the result is itself a float, stores it back into a freshly allocated box. When none of the operands are floats, the operation is emitted unchanged, exactly as before.


\section{Float Literal Support End to End}

With the heap-based boxing design fully implemented, the next task was to support float literals through the entire compiler pipeline, from instruction selection down to executable assembly. The test program \texttt{float\_literal.py} (see Appendix) served as the target: it assigns a float literal to a variable and prints it.

\subsection{Instruction Selection}

The first pass that required modifications was instruction selection. As a prerequisite, I extended the register library to include all 32 floating-point registers defined by RISC-V (see Appendix): the temporary registers \texttt{ft0}--\texttt{ft11}, the saved registers \texttt{fs0}--\texttt{fs11}, and the argument registers \texttt{fa0}--\texttt{fa7}. Under the boxed-float strategy most of these registers are not yet used, since every float value resides on the heap and is only loaded into a register for a single operation before being stored back. Nonetheless, defining the full register set lays the groundwork for future optimisations such as keeping floats in registers across operations or spilling to the stack instead of the heap.

Inside the pass itself, I threaded the types dictionary from the compiler driver down to instruction selection so that the pass could determine whether a value should be directed to a floating-point register or an integer register. Accessing a boxed float requires unboxing: loading from byte offset~7 of the tagged heap pointer to reach the payload field of the box.

\subsection{Patch Instructions for Literals}

The final pass that required changes was patch instructions, which lowers abstract move operations into concrete RISC-V assembly. To support this, I first extended the RISC-V AST with a new instruction node for floating-point moves. RISC-V provides three variants: \texttt{fmv.d.x} (integer register to float register), \texttt{fmv.x.d} (float register to integer register), and \texttt{fmv.d} (float register to float register). I defined all three in the AST.

Inside the pass, when a move instruction is encountered, I introduced a four-way case distinction based on whether the source and destination are floating-point or integer registers. Each combination selects the appropriate move instruction: \texttt{fmv.d} when both are float registers, \texttt{fmv.d.x} when only the destination is a float register, \texttt{fmv.x.d} when only the source is a float register, and the standard integer \texttt{add} (used as a move) when neither is.

\subsection{Extension of the Runtime}

To support printing floats, I added a \texttt{print\_float} function to the C runtime (\texttt{runtime.c}). This function receives a double-precision value in \texttt{fa0}---following the RISC-V calling convention for floating-point arguments---and prints it using \texttt{printf}. This mirrors the existing \texttt{print\_int64} function that the base compiler uses for integers.

\subsection{Walkthrough of the Generated Assembly}

To illustrate the end-to-end mechanics, consider how the compiler handles the statement \texttt{a: float = 3.14}. The float value is converted to its IEEE~754 bit representation as a 64-bit integer and loaded into an integer register using \texttt{li}. It is then stored at byte offset~7 from the tagged heap pointer, which places it in the payload field of the box:

\begin{lstlisting}
    li      a2, 4614253070214989087
    add     t0, zero, a2
    sd      t0, 7(a1)
\end{lstlisting}

When the program calls \texttt{print(a)}, the compiler emits instructions to unbox the float: it loads the raw bits from the heap back into an integer register, then uses \texttt{fmv.d.x} to reinterpret those bits as a double-precision value in the floating-point register \texttt{fa0}, and finally calls the runtime's \texttt{print\_float}:

\begin{lstlisting}
    ld      a1, 7(a1)
    fmv.d.x fa0, a1
    call    print_float
\end{lstlisting}

\section{Arithmetic for Floating-Point Values}

With literals and printing working, the next step was arithmetic. The base compiler only supported addition and subtraction for integers, so multiplication (\texttt{*}) and division (\texttt{/}) had to be added to the parser, type checker, and every intermediate AST regardless of floats. For floating-point values, RISC-V provides native double-precision instructions for all four operations: \texttt{fadd.d}, \texttt{fsub.d}, \texttt{fmul.d}, and \texttt{fdiv.d}.

\subsection{Transient Use of Float Registers}

The central design choice for arithmetic was to keep float registers as a transient detour rather than a place where values live. The instruction selection pass emits the following sequence for every float operation: move both unboxed operands from integer registers into \texttt{fa0} and \texttt{fa1} with \texttt{fmv.d.x}, perform the operation (e.g.\ \texttt{fadd.d fa0, fa0, fa1}), move the result back to an integer register with \texttt{fmv.x.d}, and box it on the heap immediately. Float registers are therefore never live across instruction boundaries from the register allocator's perspective.

This approach required two supporting changes. First, the heap allocation pass already introduces temporary variables for each intermediate result of a nested expression. These temporaries must be recorded in the type dictionary as \texttt{TFloat} so that downstream passes can identify them. Second, the instruction selection pass must distinguish float operations from integer ones by checking the operand types. However, the monadic normalisation pass---which flattens nested expressions into a sequence of assignments to fresh temporaries---originally did not have access to the type dictionary, leaving its temporaries untyped. To solve this, I threaded the type dictionary through the monadic pass. When the pass creates a new temporary for a non-atomic expression such as \texttt{ETupleAccess(EVar(v), 0)} where \texttt{v} is typed as \texttt{TFloat}, the temporary inherits that type. This ensures that operand types are available in instruction selection uniformly for both arithmetic and comparisons.

\subsection{Registering Float Registers in the Allocator}

Because \texttt{fa0} and \texttt{fa1} now appear explicitly in the instruction stream, the register allocator must know about them: any register that shows up in the interference graph needs a colour mapping, otherwise the graph colouring algorithm crashes when it encounters an unknown register as a neighbour of a variable. I added both registers to the blacklist at the beginning of \texttt{REG\_ORDER}, which assigns them a negative colour. This ensures the allocator recognises them but never assigns a variable to them---which is exactly the desired behaviour, since float values are boxed to the heap immediately and never need to be spilled to a float register.

\subsection{Walkthrough of the Generated Assembly}

The generated code for a float addition such as \texttt{a + b} mirrors the unbox--operate--box pattern. Both operands are loaded from the heap, moved into float registers, added, and the result is stored back:

\begin{lstlisting}
    ld      a3, 7(a3)
    fmv.d.x fa0, a2
    add     a2, zero, a3
    fmv.d.x fa1, a2
    fadd.d  fa0, fa0, fa1
    fmv.x.d a2, fa0
    add     t0, zero, a2
    sd      t0, 7(a1)
\end{lstlisting}

The first two lines unbox the first operand: \texttt{ld} loads the raw bits from the heap (at tagged offset~7), and \texttt{fmv.d.x} reinterprets them as a double in \texttt{fa0}. Lines three and four do the same for the second operand into \texttt{fa1}. The \texttt{fadd.d} performs the addition. Finally, \texttt{fmv.x.d} moves the result back to an integer register, and \texttt{sd} stores it into a freshly allocated box on the heap.


\section{Comparing Operations for Floats}

With arithmetic in place, the next step was comparisons. The RISC-V AST had to be extended with three native double-precision comparison instructions: \texttt{feq.d}, \texttt{flt.d}, and \texttt{fle.d}. As with integers, comparisons appear in two contexts that must be handled differently: assigning the result of a comparison to a variable, and branching on a comparison (e.g.\ \texttt{if}/\texttt{else}).

\subsection{Assigning Comparison Results}

When a comparison result is assigned to a variable, the compiler loads both operands, checks their types, and---if they are floats---moves them into the designated float registers to perform the comparison. The result is a~0 or~1 in an integer register, which is then tagged like any other integer and stored in the destination variable.

RISC-V only provides \texttt{feq.d} (equal), \texttt{flt.d} (less than), and \texttt{fle.d} (less or equal). The missing operators are derived with simple transformations: greater-than and greater-or-equal swap the operands (\texttt{a > b} becomes \texttt{flt.d rd, fb, fa}), while not-equal uses \texttt{feq.d} followed by an \texttt{xor} to invert the result.

\subsection{Branching on Comparison Results}

The branching case differs structurally from the integer path. For integers, RISC-V can compare and branch in a single instruction (e.g.\ \texttt{blt rs1, rs2, label}). For floats, this is not possible---no float branch instructions exist. Instead, the compiler emits the float comparison into a temporary integer register and then branches on that register with \texttt{bne} (branch if not equal to zero).

\subsection{Walkthrough of Generated Assembly}

The following walkthrough shows the branching case. As with any float operation, both values are unboxed from the heap and moved into float registers. The \texttt{flt.d} instruction compares them and writes~1 or~0 to an integer register. The result is then tagged and the branch is taken accordingly:

\begin{lstlisting}
    ld      a2, 7(a2)
    ld      a1, 7(a1)
    fmv.d.x fa0, a2
    fmv.d.x fa1, a1
    flt.d   a1, fa1, fa0
    slli    a1, a1, 1
    li      t1, 0
    bne     a1, t1, body_6
    j       orelse_7
\end{lstlisting}


\section{Type Coercion for Mixed Integer and Float Operations}

The first change required was relaxing the type checker. Previously, binary operations required both operands to have the same type. This constraint was loosened so that operands need only be individually valid for the operation---for instance, both \texttt{TInt} and \texttt{TFloat} are permitted for arithmetic---without requiring them to match. When at least one operand is a float, the result type is \texttt{TFloat}.

With the type checker permitting mixed-type expressions, the heap allocation pass handles the actual coercion. This pass already inspects each operand to determine whether it is a float and requires unboxing. When one operand is a float and the other is an integer, the pass wraps the integer in the \texttt{int\_to\_float} unary operation introduced in Section~\ref{sec:coercion}. The wrapped expression flows through the monadic and explicate passes without special handling.

In instruction selection, the \texttt{int\_to\_float} operation is lowered to a short sequence: the integer is loaded into a temporary register, untagged by shifting right by one bit (since the compiler stores integers in tagged form), and then converted to a double-precision float using the RISC-V \texttt{fcvt.d.l} instruction. The resulting float value is moved back to an integer register, ready for use in the subsequent arithmetic operation.

\subsection{Walkthrough of Generated Assembly}

The following listing shows the generated code for \texttt{3.2 * 3}, where the first operand is a float and the second is an integer. The float is unboxed from the heap as usual, while the integer literal~3 is loaded in its tagged form (value~6), untagged, and converted to a double before the multiplication:

\begin{lstlisting}
    ld      a3, 7(a3)
    li      a2, 6
    srli    a2, a2, 1
    fcvt.d.l fa0, a2
    fmv.x.d a2, fa0
    fmv.d.x fa0, a3
    fmv.d.x fa1, a2
    fmul.d  fa0, fa0, fa1
    fmv.x.d a2, fa0
\end{lstlisting}

The first line unboxes the float operand. Lines two through five handle the integer: \texttt{li} loads the tagged value~6, \texttt{srli} untags it to~3, and \texttt{fcvt.d.l} converts the integer~3 to the double~3.0 in \texttt{fa0}. The result is then moved back to an integer register with \texttt{fmv.x.d}. Finally, both operands are moved into float registers and the multiplication is performed.

\section{Extension of the Garbage Collector}

The Garbage Collector has to be extended to know whether an allocated object is a boxed float or a tuple/closure. Because I decided to encode this directly in the object's header, the implementation only required small, local changes to the compiler and the runtime, rather than a new address space or a second pointer encoding.

\subsection{Extending the Compiler}

The decision to box a float, and the code that computes and stores its value, is made earlier in the pipeline, in \texttt{pass\_5\_6\_alloc.py}: whenever an expression produces a float, that pass allocates a single-slot container for it and emits a statement that writes the computed value into that slot, exactly like writing into any other tuple field.

By the time \texttt{pass\_8\_9\_select} sees this allocation, boxing has already been decided - its job is only to choose the correct header encoding. When it lowers the allocation into a concrete header constant, it checks whether the variable being allocated is typed as a float. If it is, it sets bit~1 of the header in addition to the length that is already encoded in the upper bits; if not, bit~1 is left clear. The payload itself does not need any float-specific handling in this pass, since it is written using the same generic tuple-subscript-store logic that handles any other tuple field.

\subsection{Extending the Runtime}

The collector's sweep over to-space normally treats every word following an object's header as a potential pointer and checks it accordingly. This is exactly what would go wrong for a boxed float, since the raw bits of a double can easily satisfy that check by coincidence.

The extended sweep loop first reads bit~1 of the header. If it is set, the loop skips both the header and the single payload word without inspecting the payload at all, advancing its scan pointer by two. If it is not set, the loop proceeds as before, checking each of the object's fields individually. This one bit is enough to keep the payload of a boxed float from ever being misread as a heap pointer during collection.

\section{Generalizing the Float Box}

Prior to this extension, every place that boxed a float - a literal, the result of a unary operation, and the result of a binary operation - duplicated the same steps: check whether a collection is needed, allocate the payload, and register the resulting variable's type. I consolidated this into a single helper, \texttt{make\_float\_box(n, types)}, in \texttt{pass\_5\_6\_alloc.py}, which performs the garbage-collection check and the allocation for a box of $n$ raw words and returns both the fresh box identifier and the statements needed to construct it. \texttt{EConstFloat}, \texttt{EOp1}, and \texttt{EOp2} all now call this helper with $n=1$, rather than repeating the same allocation logic three times.

Instruction selection required no changes. The existing header-construction logic in \texttt{pass\_8\_9\_select.py} already computed \texttt{(num\_elems~<<~2)~|~0b10} generically from whatever length the allocation was given - it had simply never been called with anything other than~1.

The one change genuinely required was in the runtime. The collector's sweep, on encountering a raw object, previously skipped exactly two words unconditionally, regardless of the header's own length field. I changed this to skip \texttt{1~+~count} words, where \texttt{count} is read directly from the header, exactly as it already was for scannable objects. For $n=1$ this computes $1+1=2$, identical to the previous behaviour, so no existing test could regress; for any $n>1$ the collector now correctly treats the box as a single unit and skips the entire payload without scanning it for pointers.