\chapter{Evaluation}\label{chap:evaluation}

Development was based on 20 predesigned tests intended to ensure basic functionality and feature support, along with 3 additional tests written after completing the implementation to stress-test it and cover edge cases, on top of the 141 tests already present from the original compiler.

The practice of testing against the interpreter as the language definition was continued from earlier work: before writing any compiler code, I first defined the expected behavior via the interpreter.

One caveat did arise: the float representation of the print function differs between Python and C. Although the underlying value is the same, the C representation uses more digits after the decimal point (e.g. 3.2 vs. 3.200000000), so outputs were compared for similarity rather than exact equality, to avoid false negatives.

This also surfaced a related, known formatting difference: occasionally the two rendered outputs disagree in their final decimal digit, e.g. 3.100000 versus 3.099999 for what is otherwise the same value. The underlying IEEE 754 binary representation is identical in both languages; the difference comes from how each language's print function rounds that binary value into a decimal string. C's \texttt{printf("\%f")} always rounds to a fixed six decimal places, while Python's \texttt{print} uses the shortest decimal string that round-trips back to the exact same binary value. When the true binary value sits close to a decimal rounding boundary, these two independent rounding strategies can disagree in the last digit, even though neither is incorrect.


\section{Literals and Arithmetic}

These tests cover storing floats in variables, printing them, and performing basic arithmetic (addition, subtraction, and so on). Since this was a required feature, the implementation had to guarantee that it was fully supported.

\section{Comparison and Coercion}

This builds on the basic arithmetic tests. Comparisons required a different set of operations to be supported, and their results had to be handled differently from arithmetic results. Tests such as compare capture this by storing the result of a greater-than-or-equal comparison in a variable.

For coercion, I tested that the result has the correct representation and that operands of mixed types are converted correctly before the operation is applied.

\section{Control Flow}

These tests were targeted at making sure the extension did not break already-working features: if and while statements should not behave any differently with floats than they do with other data types, and these tests confirm that this holds.

\section{Functions and Closures}

These tests were designed to check that function return types were propagated correctly. In doing so, they uncovered a bug in which type information for a closure's captured variables and parameters was lost during closure conversion. As a result, references to captured float variables inside a closure's body were treated as plain integers further down the pipeline.

The fix was to thread the type dictionary through the closure conversion pass and to record a local type map for each generated closure function (see chapter 4 for details).

\section{Data Structures and Objects}

The main purpose of these tests was to exercise the functionality of boxed floats. Since every object, function, and data structure that has a float as a field is itself represented as a tuple, any such float had to be boxed within it.

Because this transformation was already implemented, these tests checked whether anything broke as a result of the new way float values were handled.

They also verify that boxing and unboxing inside these structures is handled correctly. The design of these tests influenced decisions such as boxing itself, since they were written before the corresponding implementation, as a way to explore how the internals should work and what needed to be supported.

In that sense, these tests directly laid the foundation for the decision to box floats.

\section{Stress tests and Edge Cases}

After finishing the core implementation, I wrote three additional test files specifically designed to test edge cases and to stress-test the garbage collector.

Even under roughly 2000 heap allocations performed in a tight loop, the garbage collector handled collection correctly and reasonably quickly. The same held for the edge-case value tests, which covered very small, very large, and signed-zero floats.

The test involving nested, chained closures revealed a real limitation: using the result of one closure call as the argument to another produced an incorrect value. The most likely cause lies somewhere in code generation or register allocation for chained closure calls, though the exact root cause was not conclusively identified within the scope of this thesis. This test was therefore moved into a separate, clearly labeled limitations folder and is discussed here rather than counted among the passing test suite.